---
title: "PA1"
author: "Bernhard Thoni"
date: "Friday, November 14, 2014"
output: html_document
---


# 0. Task: some prerequisites here (loading libraries, setting options, etc)
```{r chunk_task0, echo=T}
library(knitr)
library(lubridate)
#library(stringr)
library(ggplot2)
library(markdown)
library(data.table)
library(timeDate)
library(timeSeries)
#library(zoo)
#library(useful)
# rpubsUpload(title, htmlFile, id = NULL, properties = list(), method = getOption("rpubs.upload.method", "internal"))
opts_chunk$set(fig.path = "./figures/") # Set figures path
opts_chunk$set(scipen=1,digits=4) #just a hint from the forum
set.seed(503) #just in case! for reproducibility
# Sys.setlocale("LC_ALL", "en")
```
  

# 1. Task: Loading and preprocessing the data (incl. visual inspection)
```{r chunk_task1, echo=TRUE}
my_data<-read.csv("activity.csv",header=TRUE)   
head(my_data)
tail(my_data)
# here is preprocessing: merge date and interval (military-time) together in POSIXct, then convert to timeseries
my_data$interval_sec<-((my_data$interval %/% 100) * 3600) + ((my_data$interval %% 100) *60)
my_data$date<-as.POSIXct(my_data$date)
my_data$date_total<-my_data$date + my_data$interval_sec
my_data_p<-my_data[,c(2,1,5)]
my_ts<-as.timeSeries(my_data_p)
```
  

# 2. Task: Sumplot + Median/Mean total Number of steps taken per day, -ignoring NAs
```{r chunk_task2, echo=TRUE}
my_by<-timeSequence(from=start(my_ts), to=end(my_ts),by="day")
my_agg<-aggregate(my_ts, my_by, sum)
my_agg_m<-as.matrix(my_agg)
my_agg_df<-as.data.frame(my_agg_m)
my_agg_df$date<-rownames(my_agg_df)
qplot(x=my_agg_df$date, y=my_agg_df$steps, geom="bar", stat="identity", xlab="date", ylab="steps", main="Sum-total steps per day") #only lacks rotation of x-ticks... with ggplot2 it would be possible http://www.cookbook-r.com/Graphs/Axes_(ggplot2)/
# timeSeries::plot(my_agg$steps, type="h",col="steelblue",main="SUM-TOTAL of steps per day",xlab="date",ylab="steps") #this is just another try for plotting
my_mean<-sum(my_agg_df$steps,na.rm=T)/nrow(my_agg_df)
# I did not opt for mean(my_agg_df$steps,na.rm=T)  as this was heavily discussed in the forum!
my_median<-median(my_agg_df$steps,na.rm=T)
```

The mean total number of steps taken per day amounts to `r my_mean` steps.  
The median total number of steps taken per day amounts to `r my_median` steps.

  
# 3. Task: Finding the average daily activity pattern
```{r chunk_task3, echo=TRUE}
my_daily_act<-aggregate(steps ~ interval, data=my_data, mean)
my_daily_act$interval<-as.character.Date(my_daily_act$interval)
plot(my_daily_act$interval, my_daily_act$steps, type="l", xlab= "time of day", ylab= "steps", col="green" , lwd=2, axes=F)
axis(side=1, at=my_daily_act$interval)
axis(side=2, at=round(my_daily_act$steps,0))
title("Avg # of steps each 5 min interval")
my_max_act_time <- my_daily_act[my_daily_act$steps == max(my_daily_act$steps), 1]
my_max_act_steps <- round(my_daily_act[my_daily_act$steps == max(my_daily_act$steps), 2],0)
```

The most active 5-minute period starts at `r my_max_act_time` o'clock with `r my_max_act_steps` steps (on average).


# 4. Task: Imputation (for missing NA values)
```{r chunk_task4, echo=TRUE}
my_nr_na_TF<-is.na(my_data$steps)
my_nr_na<-sum(my_nr_na_TF)
### simple random imputation - funcion: see http://www.stat.columbia.edu/~gelman/arm/missing.pdf  , page 5
random.imp <- function (a){
    missing <- is.na(a)
    n.missing <- sum(missing)
    a.obs <- a[!missing]
    imputed <- a
    imputed[missing] <- sample (a.obs, n.missing, replace=TRUE)
    return (imputed)
}
###
my_data_imp<-my_data
my_data_imp$steps<-random.imp(my_data$steps) #let's call the function!
#
my_data_p2<-my_data_imp[,c(2,1,5)]
my_ts2<-as.timeSeries(my_data_p2)
my_by2<-timeSequence(from=start(my_ts2), to=end(my_ts2),by="day")
my_agg2<-aggregate(my_ts2, my_by2, sum)
my_agg_m2<-as.matrix(my_agg2)
my_agg_df2<-as.data.frame(my_agg_m2)
my_agg_df2$date<-rownames(my_agg_df2)
qplot(x=my_agg_df2$date, y=my_agg_df2$steps, geom="bar", stat="identity", xlab="date", ylab="steps", main="Sum-total steps per day (NAs imputed!)") #only lacks rotation of x-ticks... with ggplot2 it would be possible http://www.cookbook-r.com/Graphs/Axes_(ggplot2)/
# timeSeries::plot(my_agg$steps, type="h",col="steelblue",main="SUM-TOTAL of steps per day",xlab="date",ylab="steps") #this is just another try for plotting
my_mean2<-sum(my_agg_df2$steps,na.rm=T)/nrow(my_agg_df2)
# I did not opt for mean(my_agg_df$steps,na.rm=T)  as this was heavily discussed in the forum!
my_median2<-median(my_agg_df2$steps,na.rm=T)
```

The mean total number of steps taken per day amounts to `r my_mean2` steps.  
The median total number of steps taken per day amounts to `r my_median2` steps.  
  
Discernable difference between mean without imputation (`r my_mean`) and with imputation (`r my_mean2`)  
Discernable difference between median without imputation (`r my_median`) and with imputation (`r my_median2`)

consequently: imputation has an influence on mean and median!!!

# Task 5: Difference in Activity pattern between weekends and weekdays
```{r chunk_task5, echo=TRUE}
my_data_imp$day_of_week<-(weekdays(my_data_imp$date_total))
my_data_imp$day_of_week<-gsub("Montag","weekday",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Dienstag","weekday",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Mittwoch","weekday",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Donnerstag","weekday",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Freitag","weekday",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Samstag","weekend",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week<-gsub("Sonntag","weekend",my_data_imp$day_of_week,ignore.case=TRUE)
my_data_imp$day_of_week_fac<-factor(my_data_imp$day_of_week, labels=c("weekday","weekend"))
## preparing canvas
par(mfrow = c(2, 1))
## and now the final plot weekdays
my_daily_act2<-aggregate(steps ~ interval, data=subset(my_data_imp,day_of_week_fac=="weekday"), mean)
my_daily_act2$interval<-as.character.Date(my_daily_act2$interval)
plot(my_daily_act2$interval, my_daily_act2$steps, type="l", xlab= "time of day", ylab= "steps", col="green" , lwd=2, axes=F)
axis(side=1, at=my_daily_act2$interval)
axis(side=2, at=round(my_daily_act2$steps,0))
title("Weekdays: Avg # of steps each 5 min interval")
## and now the final plot weekends
my_daily_act3<-aggregate(steps ~ interval, data=subset(my_data_imp,day_of_week_fac=="weekend"), mean)
my_daily_act3$interval<-as.character.Date(my_daily_act3$interval)
plot(my_daily_act3$interval, my_daily_act3$steps, type="l", xlab= "time of day", ylab= "steps", col="red" , lwd=2, axes=F)
axis(side=1, at=my_daily_act3$interval)
axis(side=2, at=round(my_daily_act3$steps,0))
title("Weekends: Avg # of steps each 5 min interval")
# switching back canvas to default
par(mfrow = c(1, 1))
```




